 

### 2021-06-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a]. In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (@conditQuantifyingDeciduousnessTropical2000c, Meakem, *temperate REFS*). In this review, we focus on growing season conditions unless otherwise noted.

@@

In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained at the tops of(~100 m. tall *Sequoia sempervirens*; *Ambrose et al. 2009, 2010*).
Isolating height (>35m) from light shows that, canopy $gs$ decreases with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010]. Morphologically, leaf size [Kaare H. Jensen and Maciej A. Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] have been shown to decrease, and leaf mass per area (at greater than 20m) to increase, with height  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].

@@

#### Morphological and physiological traits

Exposed canopy leaves are better adapted to heat and water stress than shade leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*], showing greater hydraulically efficiency than shade leaves [Brodribb et al. 2010]. All these traits, collectively regulate $T_{leaf}$-photosynthetic balance. On the other hand, shaded leaves have lower convective cooling strategies, and invest more in light capture [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al. 2019]. Under homogeneous canopy light conditions, sun-leaves in different canopy orientations are primarily shaped by differences in $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].

@@

Biochemical protection against foliage light and heat damage increases with irradiance.
Caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light into heat [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. This is also important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a]. N and P are essential for photosynthetic and metabolic processes along with $VAZ$, as they contribute to enzyme formation and function, such as in Rubisco. $N_{area}$ increases in the direction of new leaf growth, where it optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018]. Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed canopies, making these species particularly advantageous [*REF*].

@@

Exposed sun-leaves, though equipped with thermoregulatory traits, are more constrained by $gs$ than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sunleaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006]. However, $T_{opt}$ of $gs$ doesnot vary between sun and shade leaves [Slot et al. 2019]. The diurnal $gs$ pattern and resulting $A_{max}$ varies between species; for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts[Bennett et al. 2015]. Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Increase in such canopy $T_{leaf}$ reduces forest GPP [@pauTropicalForestTemperature2018] in the tropics, where thermal limits are narrower than temperate forests [Perez and Feeley, 2018]. 
[**to include Teskey 2015; Breashers et al. 2021**]

@@

Warming shifts forest composition, from leaf metabolism to whole forest structure. At a leaf metabolism level, fluxnet global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]. However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosytems [@keenanWidespreadInhibitionDaytime2019]. However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Templer et al.].  Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020]. Further, combination of warm-season associated with increased VPD, along with drought exacerbates tree stress[@parkwilliamsTemperaturePotentDriver2013] leading to mortality [@gouldenCaliforniaForestDieoff2019], with implications for forest structure.

>>>>

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (@conditQuantifyingDeciduousnessTropical2000c, Meakem, *temperate REFS*). In this review, we focus on growing season conditions unless otherwise noted.

@@

In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained at the tops of(~100 m. tall *Sequoia sempervirens*; *Ambrose et al. 2009, 2010*).
Isolating height (>35m) from light shows that, canopy $gs$ decreases with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010]. Morphologically, leaf size [@jensenPhysicalLimitsLeaf2013b] and lobation [@kusiPlasticLeafMorphology2020] have been shown to decrease, and leaf mass per area (at greater than 20m) to increase, with height  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017; @marshallFoliageHeightInfluences2003].

@@

#### Anatomical, Morphological and Physiological traits

Exposed canopy environment induces traits in sun leaves that are better adaptated to heat and water stress than shade leaves through, for instance, being smaller, hairier, with greater vein density and LMA [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], as well as strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*]. These traits [see Table 1] collectively dissapate heat and regulate $T_{leaf}$-photosynthetic balance. Whereas, shade leaves have traits that enable rapid light capture in the diffused-light environment with sunflecks, for example greater SLA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013a]. In open canopies where light is comparitevely homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, than solely by light [@mediavillaFoliarPlasticityRelated2019a].

@@

Biochemical protection against foliage light and heat damage also increases with irradiance.
Caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light into heat [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. This is also important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a]. N and P are essential for photosynthetic and metabolic processes along with $VAZ$, as they contribute to enzyme formation and function, such as in Rubisco. Particularly, $N_{area}$ increases in the direction of new leaf growth, where it also optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018]. Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed canopies, making these species more advantageous [*REF*].

@@

Exposed sun-leaves, though equipped with thermoregulatory traits, are more constrained by $gs$ than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sunleaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006]. The diurnal $gs$ pattern and resulting $A_{max}$ varies between species; for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts[Bennett et al. 2015]. Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Increase in such canopy $T_{leaf}$ reduces forest GPP [@pauTropicalForestTemperature2018] in the tropics, where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].

@@

Warming shifts forest composition, from leaf metabolism to whole forest structure. At a leaf metabolism level, fluxnet global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]. However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosytems [@keenanWidespreadInhibitionDaytime2019]. However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Templer et al.].  Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020]. Further, combination of warm-season associated with increased VPD, along with drought exacerbates tree stress[@parkwilliamsTemperaturePotentDriver2013] leading to mortality [@gouldenCaliforniaForestDieoff2019], with implications for forest structure. [**to include Teskey 2015; Breashers et al. 2021**]
